---
title: "Blog Post"
author: "Scout Leonard"
date: "11/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(here)
library(lubridate)
library(tidyverse)
library(zoo)
```

# My Question:

My analysis seeks to explore the questions: Are there seasonal trends in how U.S. school food programs feed students? How did this trend change in 2020 during school closures from the Covid-19 pandemic? 

## Background

The National School Lunch Program (NSLP) is an enormous food system with enormous implications for equity in American K-12 education systems. Every day, NSLP provides ~30 million children school lunch at free or reduced prices (1). 

# Data Description

```{r}
#read in the monthly lunch data 
usda_monthly <- read.csv(here("data", "usda_monthly_data_tidy.csv"))
```

```{r}
#change first column name 
colnames(usda_monthly)[1] <- "month"

#delete the weird columns that got added between downloading the raw data to my 
#local computer and reading it to .Rmd
usda_monthly <- select(usda_monthly, -c("X", "X.1"))

#change last column name
colnames(usda_monthly)[9] <- "fiscal_year"
```

```{r}
#remove percent signs from percent_free_of_total_lunches and 
#percent_reduced_price_of_total_lunches columns
usda_monthly <- usda_monthly %>% 
 mutate(percent_free_of_total_lunches = gsub('%','', percent_free_of_total_lunches)) %>% 
  mutate(percent_reduced_price_of_total_lunches = gsub('%','', percent_reduced_price_of_total_lunches)) %>% 
  mutate(month = gsub("-", " ", month))

#convert month column to class datetime from class character
usda_monthly <- usda_monthly %>% 
  mutate("month" = zoo::as.yearmon(month, "%y %b"))
```

```{r}
#check class of all columns
lapply(usda_monthly, class)

#apply numeric to character columns
usda_monthly <- usda_monthly %>% 
  mutate_if(is.character, as.numeric)

#check to see if class conversion worked 
lapply(usda_monthly, class)
```


```{r}
#plots 

ggplot(data = usda_monthly, aes(x = month, y = total_lunches_served)) +
  geom_line()
```

# Analysis Plan

```{r}
# df = usda_monthly[, c("month", "total_lunches_served")]
# 
# df = df %>% mutate(ma = slider::slide2_dbl(.x = month, .y = total_lunches_served, .f = mean, .before = 2, .after = 2, .complete = TRUE)
```


#Summarize results visually and in words

#Next steps and future directions

#References

(1) https://www.ers.usda.gov/topics/food-nutrition-assistance/child-nutrition-programs/national-school-lunch-program/ 
